# Hurricanes and Climate Change: A Bayesian approach
#### By: Madeline Abbott, Aidan Teppema, Daisy Cho
#### December 17, 2017

```{r include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(MacBayes)
```

Loading the data...
```{r include=FALSE}
# Loading the Data
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")
hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# Recent Hurricanes--select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 
recent_hurricanes <- recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]
#Format Coordinates
dat<-as.character(recent_hurricanes$LAT)
new<-substr(dat,1,nchar(dat))
lat<-as.numeric(new)
dat<-as.character(recent_hurricanes$LONG)
new<-substr(dat,1,nchar(dat))
long<-as.numeric(new)
long<- -abs(long)
coord<-as.data.frame(long)
#Add year and name
year<-as.character(recent_hurricanes$YEAR)
name<-as.character(recent_hurricanes$NAME)
hurricane_locations<-cbind(year, name, coord, lat)

# Hurricane Frequency
# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))
hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))

# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)
# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))
# rename the columns
colnames(tempN) <- c("YEAR", "temp")
# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")
# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)
# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))
# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")
# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))
NAO_index2 <- NAO_index1[, -c(14)]
NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
NAO_index <- NAO_index3[-nrow(NAO_index3),] 
# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 
# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"
NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)
# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))
# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")
# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)

head(hurricanes_per_year_clean)
```
In the past summer, there were two major and devastating category 4 hurricanes in the US. Peoplel in Central America, Texas, Louisiana and other states lost their shelter and loved ones from Hurricane. Hurricane Irma soon after attacked as a ruining the coast of Florida. We wanted to see if this was indicative of the future and see if 

## The Datasets

To measure climate changes, we looked at four different data sets. The first data set is our main data set on hurricanes, provided by the homeland infrastructure foundation level data committee. This dataset recorded the hurricane’s wind speed, dates of occurrence, location, pressure, and the category of the hurricane. 
```{r}

```
Our indicator variables are temperature anomalies provided by NASA, CO2 provided by the national oceanic and atmospheric administration, and the North Atlantic Oscillation Index or NAOI which is provided by the national center for atmospheric research. 
The temperature anomalies are the difference from the temperature at the time from the baseline temperature, which is created from an average of 30 years of temperature data. 
NAOI is the fluctuations in the difference of atmospheric pressure at sea level (SLP) in the Northern Atlantic ocean.

## What is MCMC?

For all models we decided to run Markov chains to 

## Creating Our Models

### Hurricane Frequency

### Hurricane Windspeed

### Hurricane Duration


## Conclusion

## References

## Appendix

### Packages used

```{r echo=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(MacBayes)
```

### Data Cleaning

```{r echo=FALSE}
# Loading the Data
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")
hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# Recent Hurricanes--select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 
recent_hurricanes <- recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]
#Format Coordinates
dat<-as.character(recent_hurricanes$LAT)
new<-substr(dat,1,nchar(dat))
lat<-as.numeric(new)
dat<-as.character(recent_hurricanes$LONG)
new<-substr(dat,1,nchar(dat))
long<-as.numeric(new)
long<- -abs(long)
coord<-as.data.frame(long)
#Add year and name
year<-as.character(recent_hurricanes$YEAR)
name<-as.character(recent_hurricanes$NAME)
hurricane_locations<-cbind(year, name, coord, lat)

# Hurricane Frequency
# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))
hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))

# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)
# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))
# rename the columns
colnames(tempN) <- c("YEAR", "temp")
# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")
# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)
# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))
# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")
# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))
NAO_index2 <- NAO_index1[, -c(14)]
NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
NAO_index <- NAO_index3[-nrow(NAO_index3),] 
# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 
# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"
NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)
# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))
# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")
# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)

head(hurricanes_per_year_clean)
```

### Markov Chain for Windspeed


### Markov Chains for Frequency

#### Creating dataset for frequency, temperature, and zones

```{r}
# Adding zones to the temperature data and limiting time interval to between 1950-2008.

temp_data <- read_csv("ZonAnn.csv")
temp_data<-na.omit(temp_data) %>%  filter(1950 <= Year) %>%
  filter(Year <= 2008)
my_temp <- subset(temp_data, select = c("Year", "EQU-24N", "24N-44N", "44N-64N", "64N-90N"))
colnames(my_temp) <- c("Year", "EQUtoN24", "N24to44", "N44to64", "N64to90")
my_temp <- melt(my_temp, id=c("Year"))
colnames(my_temp) <- c("year", "zone", "temp")
```



#### Using CO2 and NAOI


#### Using Teperature and Zone cateogries



### Markov Chain for Average Duration

#### Creating a new dataset with duration, zones, and temperature

