# Hurricanes and Climate Change: A Bayesian approach
#### By: Madeline Abbott, Aidan Teppema, Daisy Cho
#### December 17, 2017

```{r echo=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(MacBayes)
```

```{r include=FALSE}
# Loading the Data
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")
hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# Recent Hurricanes--select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 
recent_hurricanes <- recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]

# Hurricane Frequency
# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))
hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))

# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)
# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))
# rename the columns
colnames(tempN) <- c("YEAR", "temp")
# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")
# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)
# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))
# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")
# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))
NAO_index2 <- NAO_index1[, -c(14)]
NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
NAO_index <- NAO_index3[-nrow(NAO_index3),] 
# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 
# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"
NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)
# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))
# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")
# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)

head(hurricanes_per_year_clean)
```

## Why Hurricanes?

In the past summer, there were two major and devastating category 4 hurricanes in the US. People in Central America, Texas, Louisiana and other states lost their shelter and loved ones from Hurricane. Hurricane Irma soon after attacked as a ruining the coast of Florida. We wanted to see if this was indicative of the future. Because of the severity of the destruction from the category 4 hurricanes we are looking at the average wind speed of hurricanes. Then, because of how quickly Hurricane Irma came after Hurricane Harvey, we decided to look at the frequency to see how many time we should expect hurricanes per year. Finally, we are looking at the average durations of the hurricanes. For example, Hurricane Harvey lasted for 17 days and we wanted to see if the durations of these devastations can be modeled. Looking at the wind speed, frequency, and durations of past hurricanes, we may be able to predict future hurricanes and be more prepared for them.

## The Datasets

To measure climate changes, we looked at four different data sets. The first data set is our main data set on hurricanes, provided by the homeland infrastructure foundation level data committee. This dataset recorded the hurricane’s wind speed, dates of occurrence, location, pressure, and the category of the hurricane. 
Our indicator variables are temperature anomalies provided by NASA, CO2 provided by the national oceanic and atmospheric administration, and the North Atlantic Oscillation Index or NAOI which is provided by the national center for atmospheric research.
C02 concentration showed fluctuation throughout the years because when plants are growing, photosynthesis outweighs respiration. As a result, plants take more CO2 out of the atmosphere during the warm months when they are growing the most. This can lead to noticeably lower CO2 concentrations in the atmosphere. Respiration occurs all the time, but dominates during the colder months of the year, resulting in higher CO2 levels in the atmosphere during those months.



The temperature anomalies are the difference from the temperature at the time from the baseline temperature, which is created from an average of 30 years of temperature data. We used temperatures because absolute temperatures, or actual temeperatures measured in current time is more prone to extraneous variables such as the temeprature measurement tower, wherease temperature anomaly are not as affected by these extraneous variables.



NAOI is the fluctuations in the difference of atmospheric pressure at sea level (SLP) in the Northern Atlantic ocean.



## What is a Markov Chain?

A Markov Chain is a chain of approximated values where one value is predicted by the value just prior to it.

## Creating Our Models

In order to answer four different research questions, we decided to used two different Bayesian modeling: A normal-normal model and Poisson regression model. A normal-normal model assumes a normal distribution, or a "bell curve", which the density curve described by its mean and standard deviation, for both the prior and the hyperpriors. A Poisson regression model, 

### Hurricane Windspeed

```{r echo=FALSE}
# Average max wind speed by year
hurricanes_wind <- group_by(recent_hurricanes, YEAR) %>%
  summarize(mean(WIND_KTS))
colnames(hurricanes_wind)[2] <- "AVG_WIND"
ggplot(hurricanes_wind, aes(x = YEAR, y = AVG_WIND)) + geom_point() + ylab("Average wind speed (kts)")
```

Using the normal-normal model, we modeled hurricane wind speed using C02 concentration and NAO index. We ran 50000 itreations of the Markov Chain and found that this model was not very good at modeling hurricane severity.






### Hurricane Frequency

```{r include=FALSE}
# Adding zones to the temperature data and limiting time interval to between 1950-2008.
temp_data<-na.omit(temp_data) %>%  filter(1950 <= Year) %>%
  filter(Year <= 2008)
my_temp <- subset(temp_data, select = c("Year", "EQU-24N", "24N-44N", "44N-64N", "64N-90N"))
colnames(my_temp) <- c("Year", "EQUtoN24", "N24to44", "N44to64", "N64to90")
my_temp <- melt(my_temp, id=c("Year"))
colnames(my_temp) <- c("year", "zone", "temp")

# Divide hurricane dataset by zones
hurricanes_zones <- hurricane_locations %>% 
  mutate(zone=cut(lat, breaks=c(0, 24, 44, 64, 90), labels=c( "EQUtoN24", "N24to44", "N44to64","N64to90"))) %>%
  group_by(year, zone)  %>%
  summarise(total = n())

# Combine the  hurricane and temperature dataset together
hurricanes_per_zone <- merge(my_temp, hurricanes_zones, by = c("year", "zone"), all.x = TRUE, all.y = TRUE)
hurricanes_per_zone[is.na(hurricanes_per_zone)] <- 0
```

```{r echo=FALSE}
# Total number of hurricanes per year
hurricanes_per_year <- group_by(recent_hurricanes, YEAR) %>%
  summarize(length(unique(NAME)))
colnames(hurricanes_per_year)[2] <- "TOTAL_HURRICANES"
ggplot(hurricanes_per_year, aes(x = YEAR, y = TOTAL_HURRICANES)) + geom_point()+xlab("Year")+ylab("Total Number of Hurricanes")+theme(axis.title = element_text(size=20))
```

In the plot above, we can see that the frequency of hurricanes generally increase throughout the years. To model the hurricane frequency per year we first used a Poisson regression model incorporating C02 concentration and NAO index. 

### Hurricane Duration

```{r echo=FALSE}
# Hurricane data only looking at Date and name
hurricane_time<-subset(recent_hurricanes, select = c(NAME,YEAR,MONTH,DAY))

#Group data by year and name and subtracting starting time from end time for each hurricane (duration)
foo <- hurricane_time %>%
  group_by(YEAR, NAME) %>%
  mutate(date = as.Date(paste(YEAR, MONTH, DAY, sep='-')), "%Y-%m-%d") %>%
  summarise(duration = max(date) - min(date))

#removed outlier
foo<-foo[!(foo$YEAR==1954 & foo$NAME=="ALICE"),]

# Averaging duration per year
foo<-group_by(foo, YEAR)%>%
  summarise(mean(duration))
# Name second column of data as Average_Duration
colnames(foo)[2] <- "duration"
colnames(foo)[1]<-"year"
#create durations with temperature data, CO2 data, NAOI data and frequency data
durations <- merge(foo, hurricanes_per_zone, by = c("year"), all.x = TRUE, all.y = TRUE)
ggplot(foo,aes(x=YEAR, y=Average_Duration))+
  geom_line()
```

In order to model average duration of hurricanes, we decided to use normal-normal model using zone categories and temperature anomalies. First, we decided to use these two variables because location seemed to have a higher correlation to hurricanes than C02 and the NAO index. Also, we decided on a normal-normal model because, the durations are not all at a constant rate depending on the hurricanes. The model seemed to be broad enough to incorporate the variability of hurricane durations.


## Conclusion

## References

* 
*
*Martyn Plummer (2016). rjags: Bayesian Graphical Models using MCMC. R package version 4-6.
  https://CRAN.R-project.org/package=rjags

## Appendix

### Packages used

```{r eval=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(MacBayes)
```

### Data Cleaning

```{r eval=FALSE}
# Loading the Data
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")
hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# Recent Hurricanes--select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 
recent_hurricanes <- recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]

# Hurricane Frequency
# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))
hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))

# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)
# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))
# rename the columns
colnames(tempN) <- c("YEAR", "temp")
# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")
# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)
# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))
# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")
# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))
NAO_index2 <- NAO_index1[, -c(14)]
NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
NAO_index <- NAO_index3[-nrow(NAO_index3),] 
# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 
# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"
NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)
# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))
# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")
# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)

head(hurricanes_per_year_clean)
```

### Markov Chain for Windspeed

```{r eval=FALSE}
# specify the model
windSpeed_model <- "model{
    #Data
    for(i in 1:length(y)) {
        y[i] ~ dnorm(beta0 + beta1*x1[i] + beta2*x2[i], tau)

    }
    #Priors
    beta0 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta1 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta2 ~ dnorm(0, 1/(1000)^2) #PRECISION
    tau ~ dgamma(0.001, 0.001)

}"

#*set up an algorithm to simulate the posterior by
#*combining the model (games_model) and data (x)
#*set the random number seed
windSpeed_jags <- jags.model(textConnection(windSpeed_model),data=list(y=HURRICANES$avg_windspeed,x1=HURRICANES$avg_CO2,x2=HURRICANES$avg_NAO), inits = list(beta0 = 5, beta1 = 0, beta2 = 0))

#*simulate a sample from the posterior
#*note that we specify both mu and tau variables
windSpeed_sim <- coda.samples(windSpeed_jags, variable.names=c("beta0","beta1","beta2"), n.iter=500000)


#*store the samples in a data frame:
windSpeed_sample <- data.frame(step=1:500000, windSpeed_sim[[1]])
```

### Markov Chains for Frequency

#### Creating dataset for frequency, temperature, and zones

```{r eval=FALSE}
# Adding zones to the temperature data and limiting time interval to between 1950-2008.
temp_data<-na.omit(temp_data) %>%  filter(1950 <= Year) %>%
  filter(Year <= 2008)
my_temp <- subset(temp_data, select = c("Year", "EQU-24N", "24N-44N", "44N-64N", "64N-90N"))
colnames(my_temp) <- c("Year", "EQUtoN24", "N24to44", "N44to64", "N64to90")
my_temp <- melt(my_temp, id=c("Year"))
colnames(my_temp) <- c("year", "zone", "temp")

# Divide hurricane dataset by zones
hurricanes_zones <- hurricane_locations %>% 
  mutate(zone=cut(lat, breaks=c(0, 24, 44, 64, 90), labels=c( "EQUtoN24", "N24to44", "N44to64","N64to90"))) %>%
  group_by(year, zone)  %>%
  summarise(total = n())

# Combine the  hurricane and temperature dataset together
hurricanes_per_zone <- merge(my_temp, hurricanes_zones, by = c("year", "zone"), all.x = TRUE, all.y = TRUE)
hurricanes_per_zone[is.na(hurricanes_per_zone)] <- 0
```

#### Using CO2 and NAOI



#### Using Teperature and Zone cateogories

```{r eval=FALSE}
hur_mod3 <- " model {
  for (i in 1:length(total)) {
      total[i] ~ dpois(lam[i])
      log(lam[i]) = beta0 + beta1*X1[i] + beta2[X2[i]]
  }

  beta0 ~ dnorm(0.0, 1.0/1e4)
  beta1 ~ dnorm(0.0, 1.0/1e4)
  beta2[1] <- 0
  for (i in 2:4) {
    beta2[i] ~ dnorm(0.0, 1.0/1e4)
  }
} "

# set up an algorithm to simulate the posterior by combining the model and data (x)
# set the random number seed
freq_jags3 <- jags.model(textConnection(hur_mod3),data=list(total = hurricanes_per_zone$total, X1 = hurricanes_per_zone$temp, X2 = hurricanes_per_zone$zone), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=2000))

# simulate a sample from the posterior
freq_sim3 <- coda.samples(freq_jags3, variable.names = c("beta1", "beta2"), n.iter=1000)

# store the samples in a data frame:
freq_sample3 <- data.frame(step = 1:1000, freq_sim3[[1]])
head(freq_sample3, 10)
plot(freq_sim3)
```

### Markov Chain for Average Duration

#### Creating a new dataset with duration, zones, and temperature

```{r eval=FALSE}
# Hurricane data only looking at Date and name
hurricane_time<-subset(recent_hurricanes, select = c(NAME,YEAR,MONTH,DAY))

#Group data by year and name and subtracting starting time from end time for each hurricane (duration)
foo <- hurricane_time %>%
  group_by(YEAR, NAME) %>%
  mutate(date = as.Date(paste(YEAR, MONTH, DAY, sep='-')), "%Y-%m-%d") %>%
  summarise(duration = max(date) - min(date))

#removed outlier
foo<-foo[!(foo$YEAR==1954 & foo$NAME=="ALICE"),]

# Averaging duration per year
foo<-group_by(foo, YEAR)%>%
  summarise(mean(duration))
# Name second column of data as Average_Duration
colnames(foo)[2] <- "duration"
colnames(foo)[1]<-"year"
#create durations with temperature data, CO2 data, NAOI data and frequency data
durations <- merge(foo, hurricanes_per_zone, by = c("year"), all.x = TRUE, all.y = TRUE)
```

#### Running the model

```{r eval=FALSE}
library(rjags)

#specify the model
duration_model <- "model{
    #Data
    for(i in 1:length(y)) {
        y[i] ~ dnorm(beta0 + beta1*X1[i] + beta2[X2[i]],tau)
    }

    #Priors
    beta0 ~ dnorm(0.0, 1.0/1e4)
    beta1 ~ dnorm(0.0, 1.0/1e4)
    beta2[1] <- 0
    for (i in 2:4) {
      beta2[i] ~ dnorm(0.0, 1.0/1e4)
  }
    tau ~ dgamma(.001, .001)
}"

duration_jags <- jags.model(textConnection(duration_model), data=list(y = durations$duration, X1 = durations$temp, X2 = durations$zone), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

duration_sim <- coda.samples(duration_jags, variable.names=c("beta0","beta1", "beta2","tau"), n.iter=10000)

duration_samples <- data.frame(duration_sim[[1]])
```