# Hurricanes and Climate Change: A Bayesian approach
#### By: Madeline Abbott, Aidan Teppema, Daisy Cho
#### December 17, 2017

```{r include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(MacBayes)
```

Loading the data...
```{r include=FALSE}
# Loading the Data
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")
hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# Recent Hurricanes--select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 
recent_hurricanes <- recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]
#Format Coordinates
dat<-as.character(recent_hurricanes$LAT)
new<-substr(dat,1,nchar(dat))
lat<-as.numeric(new)
dat<-as.character(recent_hurricanes$LONG)
new<-substr(dat,1,nchar(dat))
long<-as.numeric(new)
long<- -abs(long)
coord<-as.data.frame(long)
#Add year and name
year<-as.character(recent_hurricanes$YEAR)
name<-as.character(recent_hurricanes$NAME)
hurricane_locations<-cbind(year, name, coord, lat)

# Hurricane Frequency
# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))
hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))

# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)
# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))
# rename the columns
colnames(tempN) <- c("YEAR", "temp")
# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")
# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)
# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))
# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")
# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))
NAO_index2 <- NAO_index1[, -c(14)]
NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
NAO_index <- NAO_index3[-nrow(NAO_index3),] 
# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 
# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"
NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)
# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))
# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")
# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)

# Wind speed data
# Create new data set with windspeed and year
windspeed_yr <- hurricane_data %>%
  group_by(YEAR) %>%
  summarise(avg_windspeed = mean(WIND_KTS))

# Merge this with the other data set
hurricanes_per_year_clean <- merge(x = hurricanes_per_year_clean, y = windspeed_yr, by = "YEAR", all.x = TRUE)

head(hurricanes_per_year_clean)

head(hurricanes_per_year_clean)
```
In the past summer, there were two major and devastating category 4 hurricanes in the US. People in Central America, Texas, Louisiana and other states lost security and loved ones from the hurricanes. Hurricane Irma soon after attacked as a ruining the coast of Florida. 

We wanted to see if this increased number and severity of hurricanes are indicative of a trend and if we could predict if this trend would continue into the future.

## The Datasets

To measure climate changes, we looked at four different data sets. The first data set is our main data set on hurricanes, provided by the homeland infrastructure foundation level data committee. This dataset recorded the hurricane’s wind speed, dates of occurrence, location, pressure, and the category of the hurricane. 
```{r}

```
Our indicator variables are temperature anomalies provided by NASA, CO2 provided by the national oceanic and atmospheric administration, and the North Atlantic Oscillation Index or NAOI which is provided by the national center for atmospheric research. 
The temperature anomalies are the difference from the temperature at the time from the baseline temperature, which is created from an average of 30 years of temperature data. 
NAOI is the fluctuations in the difference of atmospheric pressure at sea level (SLP) in the Northern Atlantic ocean.

## What is MCMC?

For all our models we ran Markov chains in order to simulate the posterior results after running our models in rjags.

## Creating Our Models

### Hurricane Windspeed

For our first model, we tried to model hurricane severity over time using the NAO index and average atmospheric CO2 levels (we removed average ocean temperature from our model because it was collinear with CO2). The variable we used to indicate hurricane severity was wind speed, since the highest windspeed of a hurricane is what is used to categorize it.

We used a normal-normal model to attempt to model wind speed with CO2 and NAO over time. The normal-normal model is called that because both the prior and posterior distributions of the data are normally distributed. All of the priors for this model are vague because, prior to creating and running our models, we didn't know much about the impact of NAO or CO2 on hurricane wind speed.

### Hurricane Frequency

Our second and third models attempt to show a relationship between CO2, NAO, and the frequency of hurricanes over time. 

### Hurricane Duration

Our final model attempted to model the duration of the hurricane w

## Conclusion



## References

## Appendix

### Packages used

```{r eval=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(MacBayes)
```

### Data Cleaning

```{r eval=FALSE}
# Loading the Data
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")
hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# Recent Hurricanes--select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 
recent_hurricanes <- recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]

# Hurricane Frequency
# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))
hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))

# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)
# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))
# rename the columns
colnames(tempN) <- c("YEAR", "temp")
# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")
# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)
# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))
# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")
# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))
NAO_index2 <- NAO_index1[, -c(14)]
NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
NAO_index <- NAO_index3[-nrow(NAO_index3),] 
# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 
# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"
NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)
# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))
# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")
# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)

# Wind speed data
# Create new data set with windspeed and year
windspeed_yr <- hurricane_data %>%
  group_by(YEAR) %>%
  summarise(avg_windspeed = mean(WIND_KTS))

# Merge this with the other data set
hurricanes_per_year_clean <- merge(x = hurricanes_per_year_clean, y = windspeed_yr, by = "YEAR", all.x = TRUE)

head(hurricanes_per_year_clean)
```

### Markov Chain for Windspeed


### Markov Chains for Frequency

#### Creating dataset for frequency, temperature, and zones

```{r eval=FALSE}
# Adding zones to the temperature data and limiting time interval to between 1950-2008.
temp_data <- read_csv("ZonAnn.csv")
temp_data<-na.omit(temp_data) %>%  filter(1950 <= Year) %>%
  filter(Year <= 2008)
my_temp <- subset(temp_data, select = c("Year", "EQU-24N", "24N-44N", "44N-64N", "64N-90N"))
colnames(my_temp) <- c("Year", "EQUtoN24", "N24to44", "N44to64", "N64to90")
my_temp <- melt(my_temp, id=c("Year"))
colnames(my_temp) <- c("year", "zone", "temp")

# Divide hurricane dataset by zones
hurricanes_zones <- hurricane_locations %>% 
  mutate(zone=cut(lat, breaks=c(0, 24, 44, 64, 90), labels=c( "EQUtoN24", "N24to44", "N44to64","N64to90"))) %>%
  group_by(year, zone)  %>%
  summarise(total = n())

# Combine the  hurricane and temperature dataset together
hurricanes_per_zone <- merge(my_temp, hurricanes_zones, by = c("year", "zone"), all.x = TRUE, all.y = TRUE)
hurricanes_per_zone[is.na(hurricanes_per_zone)] <- 0
```

```{r}

#head(hurricanes_per_zone)
#hurricanes_per_zone$zone
#EQUtoN24 N24to44  N44to64  N64to90

```

#### Using CO2 and NAOI



#### Using Teperature and Zone cateogories

```{r eval=FALSE}
hur_mod3 <- " model {
  for (i in 1:length(total)) {
      total[i] ~ dpois(lam[i])
      log(lam[i]) = beta0 + beta1*X1[i] + beta2[X2[i]]
  }

  beta0 ~ dnorm(0.0, 1.0/1e4)
  beta1 ~ dnorm(0.0, 1.0/1e4)
  beta2[1] <- 0
  for (i in 2:4) {
    beta2[i] ~ dnorm(0.0, 1.0/1e4)
  }
} "

# set up an algorithm to simulate the posterior by combining the model and data (x)
# set the random number seed
freq_jags3 <- jags.model(textConnection(hur_mod3),data=list(total = hurricanes_per_zone$total, X1 = hurricanes_per_zone$temp, X2 = hurricanes_per_zone$zone), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=2000))

# simulate a sample from the posterior
freq_sim3 <- coda.samples(freq_jags3, variable.names = c("beta1", "beta2"), n.iter=1000)

# store the samples in a data frame:
freq_sample3 <- data.frame(step = 1:1000, freq_sim3[[1]])
head(freq_sample3, 10)
plot(freq_sim3)
```

### Markov Chain for Average Duration

#### Creating a new dataset with duration, zones, and temperature

```{r eval=FALSE}
# Hurricane data only looking at Date and name
hurricane_time<-subset(recent_hurricanes, select = c(NAME,YEAR,MONTH,DAY))

#Group data by year and name and subtracting starting time from end time for each hurricane (duration)
foo <- hurricane_time %>%
  group_by(YEAR, NAME) %>%
  mutate(date = as.Date(paste(YEAR, MONTH, DAY, sep='-')), "%Y-%m-%d") %>%
  summarise(duration = max(date) - min(date))

#removed outlier
foo<-foo[!(foo$YEAR==1954 & foo$NAME=="ALICE"),]

# Averaging duration per year
foo<-group_by(foo, YEAR)%>%
  summarise(mean(duration))
# Name second column of data as Average_Duration
colnames(foo)[2] <- "duration"
colnames(foo)[1]<-"year"
#create durations with temperature data, CO2 data, NAOI data and frequency data
durations <- merge(foo, hurricanes_per_zone, by = c("year"), all.x = TRUE, all.y = TRUE)

```

#### Running the model

```{r eval=FALSE}
library(rjags)

#specify the model
duration_model <- "model{
    #Data
    for(i in 1:length(y)) {
        y[i] ~ dnorm(beta0 + beta1*X1[i] + beta2[X2[i]],tau)
    }

    #Priors
    beta0 ~ dnorm(0.0, 1.0/1e4)
    beta1 ~ dnorm(0.0, 1.0/1e4)
    beta2[1] <- 0
    for (i in 2:4) {
      beta2[i] ~ dnorm(0.0, 1.0/1e4)
  }
    tau ~ dgamma(.001, .001)
}"

duration_jags <- jags.model(textConnection(duration_model), data=list(y = durations$duration, X1 = durations$temp, X2 = durations$zone), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

duration_sim <- coda.samples(duration_jags, variable.names=c("beta0","beta1", "beta2","tau"), n.iter=10000)

duration_samples <- data.frame(duration_sim[[1]])
```