# Project Checkpoint 6

**NAMES:**  Madeline Abbott, Aidan Teppema, Daisy Cho

\
\
\
\


```{r include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
```

```{r include=FALSE}
# Loading the Data

# Hurricane Frequency
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")

hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 
recent_hurricanes <- recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]

# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))

hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))

#Format Coordinates
#dat<-as.character(recent_hurricanes$LAT)
#new<-substr(dat,1,nchar(dat))
#lat<-as.numeric(new)
#dat<-as.character(recent_hurricanes$LONG)
#new<-substr(dat,1,nchar(dat))
#long<-as.numeric(new)
#long<- -abs(long)
#coord<-as.data.frame(long)
#Add year and name
#year<-as.character(recent_hurricanes$YEAR)
#name<-as.character(recent_hurricanes$NAME)
#coord<-cbind(year, name, coord, lat)

```

Sub-data
```{r}
# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)

# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))

# rename the columns
colnames(tempN) <- c("YEAR", "temp")

# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")

# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)

# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))

# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")

# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))

NAO_index2 <- NAO_index1[, -c(14)]

NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
  
NAO_index <- NAO_index3[-nrow(NAO_index3),] 

# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 

# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"

NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)

# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))

# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")

# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)
```

# Progress Made:


# Group Member Roles:

* Madeline--
* Daisy--
* Aidan--


# Modeling:

\



### Question 1.

Can we model the change in frequency of hurricanes based on trends in hurricane windspeed and category, temperature, CO2, and NAO index?


Plot distribution of hurricanes per year
```{r}
ggplot(hurricanes_per_year, aes(x = YEAR, y = TOTAL_H)) + geom_point(color = "orange")
```

Plot hurricanes per year by temperature
```{r}
ggplot(hurricanes_per_year_clean, aes(x = temp, y = TOTAL_H)) + geom_point(color = "skyblue")
```

Hurricanes per year by CO2 levels
```{r}
ggplot(hurricanes_per_year_clean, aes(x = avg_CO2, y = TOTAL_H)) + geom_point(color = "darkslateblue")
```

Hurricanes per year by CO2 levels
```{r}
ggplot(hurricanes_per_year_clean, aes(x = avg_NAO, y = TOTAL_H)) + geom_point(color = "darkseagreen")
```


### Model 1: 

Let:

$Y_{t} =$ total hurricanes in year $t$

$X_{1t} =$ average yearly temperature in year $t$

$X_{2t} =$ yearly CO2 concentration in year $t$

$X_{3t} =$ NAO index in year $t$

We propose the following model:

$$Y_t | \beta_{0}, \beta_{1}, \beta_{2}, \beta_{3} \sim N(\beta_0 + \beta_{1}X_{1t} + \beta_{2}X_{2t} + \beta_{3}X_{3t}, \theta^2)$$

$$\beta_0 \sim N(0, 1000^2)$$

$$\beta_{1} \sim N(0, 1000^2)$$

$$\beta_{2} \sim N(0, 1000^2)$$

$$\beta_{3} \sim N(0, 1000^2)$$

$$(\theta^2)^{-1} \sim Gamma(0.001, 0.001)$$


```{r}
# specify the model
hurricane_freq_model1 <- "model{
  # Data
  for (t in 1:length(y)) {
    y[t] ~ dnorm(beta0 + beta1*x1[t] + beta2*x2[t] + beta3*x3[t], tau) #PRECISION
    }

  #Priors
  beta0 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta1 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta2 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta3 ~ dnorm(0, 1/(1000)^2) #PRECISION
  tau ~ dgamma(0.001, 0.001)
}"


# set up an algorithm to simulate the posterior by combining the model (hurricane_freq_model1) and data (y)
# set the random number seed
hurricane_freq_jags1 <- jags.model(textConnection(hurricane_freq_model1), data=list(y=hurricanes_per_year_clean$TOTAL_H, x1=hurricanes_per_year_clean$temp, x2=hurricanes_per_year_clean$avg_CO2, x3=hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

# simulate a sample from the posterior
# note that we specify both mu and tau variables
hurricane_freq_sim1 <- coda.samples(hurricane_freq_jags1, variable.names = c("beta0", "beta1", "beta2", "beta3", "tau"), n.iter=100)

# store the samples in a data frame:
hurricane_freq_sample1 <- data.frame(step = 1:100, hurricane_freq_sim1[[1]])
head(hurricane_freq_sample1)

summary(hurricane_freq_sample1)
#plot(hurricane_freq_sample1)
```



### Model 2:

Let:

$Y_{t} =$ total hurricanes in year $t$

$X_{1t} =$ average yearly temperature in year $t$

$X_{2t} =$ yearly CO2 concentration in year $t$

$X_{3t} =$ NAO index in year $t$

We propose the following model:

$$Y_t | \beta_{0}, \beta_{1}, \beta_{2}, \beta_{3} \sim N(\beta_0 + \beta_{1}X_{1t} + \beta_{2}X_{2t} + \beta_{3}X_{3t}, \theta^2)$$

$$\beta_0 \sim N(\mu_{0}, 1000^2)$$

$$\beta_{1t} \sim N(\mu_{1}, 1000^2)$$

$$\beta_{2t} \sim N(\mu_{2}, 1000^2)$$

$$\beta_{3t} \sim N(\mu_{3}, 1000^2)$$

$$(\theta^2)^{-1} \sim Gamma(0.001, 0.001)$$

$$\mu_{0} \sim N(0, 1000^2)$$

$$\mu_{1} \sim N(0, 1000^2)$$

$$\mu_{2} \sim N(0, 1000^2)$$

$$\mu_{3} \sim N(0, 1000^2)$$

```{r}
# specify the model
hurricane_freq_model2 <- "model{
  # Data
  for (t in 1:length(y)) {
    y[t] ~ dnorm(beta0 + beta1*x1[t] + beta2*x2[t] + beta3*x3[t], tau) #PRECISION
    }

  #Priors
  beta0 ~ dnorm(mu0, 1/(1000)^2) #PRECISION
  beta1 ~ dnorm(mu1, 1/(1000)^2) #PRECISION
  beta2 ~ dnorm(mu2, 1/(1000)^2) #PRECISION
  beta3 ~ dnorm(mu3, 1/(1000)^2) #PRECISION
  tau ~ dgamma(0.001, 0.001)
  mu0 ~ dnorm(0, 1/(1000)^2)
  mu1 ~ dnorm(0, 1/(1000)^2)
  mu2 ~ dnorm(0, 1/(1000)^2)
  mu3 ~ dnorm(0, 1/(1000)^2)
}"


# set up an algorithm to simulate the posterior by combining the model (hurricane_freq_model2) and data (y)
# set the random number seed
hurricane_freq_jags2 <- jags.model(textConnection(hurricane_freq_model2), data=list(y=hurricanes_per_year_clean$TOTAL_H, x1=hurricanes_per_year_clean$temp, x2=hurricanes_per_year_clean$avg_CO2, x3=hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

# simulate a sample from the posterior
# note that we specify both mu and tau variables
hurricane_freq_sim2 <- coda.samples(hurricane_freq_jags2, variable.names = c("beta0", "beta1", "beta2", "beta3", "tau"), n.iter=100)

# store the samples in a data frame:
hurricane_freq_sample2 <- data.frame(step = 1:100, hurricane_freq_sim2[[1]])
head(hurricane_freq_sample2)

summary(hurricane_freq_sample2)
#plot(hurricane_freq_sample1)
```

### Question 2.

Can we model  the trend of average durations of hurricanes over the years?

Model 1:

Prior: 
$$X \sim Norm(\mu,\tau)$$

Hyperpriors:
$$\mu \sim N(0, 1000)$$

$$\tau \sim \Gamma(2, 0.05)$$

Model 2:

Prior: $$X \sim N(mu, tau)$$

Given length(x)= number of years
$$\mu\sim N(6.37, 10^2)$$

$$\tau \sim \Gamma(9, 6)$$

Create new dataset on average durations per year
```{r echo=FALSE}
# Hurricane data only looking at Date and name
hurricane_time<-subset(hurricanes_recent, select = c(NAME,YEAR,MONTH,DAY))

#Group data by year and name and subtracting starting time from end time for each hurricane (duration)
foo <- hurricane_time %>%
  group_by(YEAR, NAME) %>%
  mutate(date = as.Date(paste(YEAR, MONTH, DAY, sep='-')), "%Y-%m-%d") %>%
  summarise(duration = max(date) - min(date))

#removed outlier
foo<-foo[!(foo$YEAR==1954 & foo$NAME=="ALICE"|foo$YEAR==2005),]

# Averaging duration per year
foo1<-group_by(foo, YEAR)%>%
  summarise(mean(duration))
colnames(foo1)[2] <- "Average_Duration"
ggplot(foo1, aes(x=Average_Duration))+
  geom_histogram(boundary=0,color="white", aes(y=..density..))
ggplot(foo1,aes(x=YEAR, y=Average_Duration))+
  geom_line()
```

### Model 1:

```{r}
library(rjags)

#specify the model
duration_model <- "model{
    #Data
    for(i in 1:length(x)) {
        y[i] ~ dgamma(a[i],b)
        a[i] ~ dpois(lambda[i])
    }

    #Priors
    b ~ dgamma(1,1)
    beta0 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta1 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta2 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta3 ~ dnorm(0, 1/(1000)^2) #PRECISION
    tau ~ dgamma(0.001, 0.001)
}"



duration_jags <- jags.model(textConnection(duration_model), data=list(y=foo_train$Average_Duration, x1=hurricanes_per_year_clean$temp, x2=hurricanes_per_year_clean$avg_CO2, x3=hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))



duration_sim <- coda.samples(duration_jags, variable.names=c("b","beta0","beta1","beta2","beta3", "tau"), n.iter=10000)

duration_samples <- data.frame(step=1:10000, duration_sim[[1]])
head(duration_samples)
plot(duration_sim)
summary(duration_sim)
```

Model 2:

Create training and testing data
```{r message=FALSE, warning=FALSE}
data(foo)

## 50% of the sample size
smp_size <- floor(0.50 * nrow(foo))

## set the seed to make your partition reproductible
set.seed(123)
train_ind <- sample(seq_len(nrow(foo)), size = smp_size)

foo_train <- foo[train_ind, ]
foo_test <- foo[-train_ind, ]
```

Create model using training data
```{r}
library(rjags)

#specify the model
duration2_model <- "model{
    #Data
    for(i in 1:length(x)) {
        y[i] ~ dnorm(beta0 +beta1*x[i],tau)
    }

    #Priors
    beta0 ~ dnorm(6,1/1000)
    beta1~dnorm(0,1/1000)
    tau ~ dgamma(9, 6)
}"



duration2_jags <- jags.model(textConnection(duration2_model), data=list(x=foo_train$Average_Duration), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))



duration2_sim <- coda.samples(duration2_jags, variable.names=c("beta0","beta1","tau"), n.iter=10000)

duration2_samples <- data.frame(step=1:10000, duration_sim[[1]])
head(duration2_samples)
plot(duration2_sim)
summary(duration2_sim)
```


Test model on testing data
```{r}
library(rjags)

#specify the model
duration3_model <- "model{
    #Data
    for(i in 1:length(x)) {
        y[i] ~ dnorm(beta0 +beta1*x[i],tau)
    }

    #Priors
    beta0 ~ dnorm(6,1/1000)
    beta1~dnorm(0,1/1000)
    tau ~ dgamma(9, 6)
}"



duration3_jags <- jags.model(textConnection(duration3_model), data=list(x=foo_test$Average_Duration), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))



duration3_sim <- coda.samples(duration3_jags, variable.names=c("beta0","beta1","tau"), n.iter=10000)

duration3_samples <- data.frame(step=1:10000, duration_sim[[1]])
head(duration3_samples)
plot(duration3_sim)
summary(duration3_sim)
```

### Questions 3.




