# Project Checkpoint 6

**NAMES:**  Madeline Abbott, Aidan Teppema, Daisy Cho

\
\
\
\


```{r include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(MacBayes)
```

Loading the data...
```{r include=FALSE}
# Loading the Data
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")
hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# Recent Hurricanes--select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 
recent_hurricanes <- recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]
<<<<<<< HEAD

# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))

hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))

=======
>>>>>>> a22f559e9cd25af3504a9da11315b346e4d5b40e
#Format Coordinates
#dat<-as.character(recent_hurricanes$LAT)
#new<-substr(dat,1,nchar(dat))
#lat<-as.numeric(new)
#dat<-as.character(recent_hurricanes$LONG)
#new<-substr(dat,1,nchar(dat))
#long<-as.numeric(new)
#long<- -abs(long)
#coord<-as.data.frame(long)
#Add year and name
#year<-as.character(recent_hurricanes$YEAR)
#name<-as.character(recent_hurricanes$NAME)
#coord<-cbind(year, name, coord, lat)

<<<<<<< HEAD
```
=======
# Hurricane Frequency
# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))
hurricanes_per_basin <- recent_hurricanes %>%
  group_by(BASIN) %>%
  summarise(TOTAL_H = n_distinct(NAME, DAY))
>>>>>>> a22f559e9cd25af3504a9da11315b346e4d5b40e

Sub-data
```{r}
# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)
# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))
# rename the columns
colnames(tempN) <- c("YEAR", "temp")
# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")
# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)
# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))
# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")
# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))
NAO_index2 <- NAO_index1[, -c(14)]
NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
NAO_index <- NAO_index3[-nrow(NAO_index3),] 
# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 
# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"
NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)
# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))
# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")
# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)
```



# Progress Made:

This week, we worked to fix the model that was not functioning last week (modeling windspeed).  We created a new model for annual hurricane rate using poisson regression.  We summarized this new model (and also last week's normal-normal model of hurricane rate) with running mean plots and found the credible intervals of the parameters.

# Group Member Roles:

* Madeline--summarized normal-normal model results and worked on poisson regression for hurricane rate.
* Daisy--worked on modeling duration of hurricanes, tried to create new dataset of hurricane landfall
* Aidan--corrected, expanded, and summarized the windspeed model


# Modeling:

\

Things to think about:

* limit data to hurricanes in the Atlantic

* include year as a predictor in models


## Question 1.

Can we model the change in frequency of hurricanes based on trends in hurricane windspeed and category, temperature, CO2, and NAO index?


\

###Some plots...

\

Plot histogram of annual hurricane frequency
```{r}
# hurricanes
ggplot(hurricanes_per_year_clean, aes(x = TOTAL_H)) + geom_histogram(color = 'white', fill = 'goldenrod1', aes(y = ..density..))

# temp
ggplot(hurricanes_per_year_clean, aes(x = temp)) + geom_histogram(color = 'white', fill = 'coral', aes(y = ..density..))

#co2
ggplot(hurricanes_per_year_clean, aes(x = avg_CO2)) + geom_histogram(color = 'white', fill = 'darkorchid2', aes(y = ..density..))

# NAO
ggplot(hurricanes_per_year_clean, aes(x = avg_NAO)) + geom_histogram(color = 'white', fill = 'darkseagreen', aes(y = ..density..))
```


Plot hurricanes per year by temperature
```{r}
ggplot(hurricanes_per_year_clean, aes(x = temp, y = TOTAL_H)) + geom_point(color = "firebrick1") + xlab("Temperature") + ylab("Total Number of Hurricanes")
```

Hurricanes per year by CO2 levels
```{r}
ggplot(hurricanes_per_year_clean, aes(x = avg_CO2, y = TOTAL_H)) + geom_point(color = "darkslateblue") + xlab("CO2 Concentration") + ylab("Total Number of Hurricanes")
```

Hurricanes per year by NAO levels
```{r}
ggplot(hurricanes_per_year_clean, aes(x = avg_NAO, y = TOTAL_H)) + geom_point(color = "aquamarine4") + xlab("NAO Index") + ylab("Total Number of Hurricanes")
```


### Model 1: Normal-normal

Question 1:Can we model the change in frequency of hurricanes based on trends in hurricane windspeed and category, temperature, CO2, and NAO index?

\

Let:

$Y_{t} =$ total hurricanes in year $t$

$X_{1t} =$ average yearly temperature in year $t$

$X_{2t} =$ yearly CO2 concentration in year $t$

$X_{3t} =$ NAO index in year $t$

We propose the following model:

$$Y_t | \beta_{0}, \beta_{1}, \beta_{2}, \beta_{3} \sim N(\beta_0 + \beta_{1}X_{1t} + \beta_{2}X_{2t} + \beta_{3}X_{3t}, \theta^2)$$

$$\beta_0 \sim N(0, 1000^2)$$

$$\beta_{1} \sim N(0, 1000^2)$$

$$\beta_{2} \sim N(0, 1000^2)$$

$$\beta_{3} \sim N(0, 1000^2)$$

$$(\theta^2)^{-1} \sim Gamma(0.001, 0.001)$$


Build the model--normal-normal
```{r}

# specify the model
hurricane_freq_model1 <- "model{
  # Data
  for (t in 1:length(y)) {
    y[t] ~ dnorm(beta0 + beta1*x1[t] + beta2*x2[t] + beta3*x3[t], tau) #PRECISION
    }

  #Priors
  beta0 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta1 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta2 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta3 ~ dnorm(0, 1/(1000)^2) #PRECISION
  tau ~ dgamma(0.001, 0.001)
}"


# set up an algorithm to simulate the posterior by combining the model (hurricane_freq_model1) and data (y)
# set the random number seed
hurricane_freq_jags1 <- jags.model(textConnection(hurricane_freq_model1), data=list(y=hurricanes_per_year_clean$TOTAL_H, x1=hurricanes_per_year_clean$temp, x2=hurricanes_per_year_clean$avg_CO2, x3=hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=2000))

# simulate a sample from the posterior
# note that we specify both mu and tau variables
hurricane_freq_sim1 <- coda.samples(hurricane_freq_jags1, variable.names = c("beta0", "beta1", "beta2", "beta3", "tau"), n.iter=50000)

# store the samples in a data frame:
hurricane_freq_sample1 <- data.frame(step = 1:50000, hurricane_freq_sim1[[1]])
head(hurricane_freq_sample1)

#summary(hurricane_freq_sim1)
#plot(hurricane_freq_sim1)
#summary(hurricane_freq_sample1)
#plot(hurricane_freq_sample1)

```

Do parameter estimations converge?
```{r}
running_mean_plot(x = hurricane_freq_sample1$beta1, se = TRUE)
running_mean_plot(x = hurricane_freq_sample1$beta2, se = TRUE)
running_mean_plot(x = hurricane_freq_sample1$beta3, se = TRUE)
```

Are the predictor variables useful?  Look at model parameters
```{r}
sum(hurricane_freq_sample1$beta0)/10000
sum(hurricane_freq_sample1$beta1)/10000
sum(hurricane_freq_sample1$beta2)/10000
sum(hurricane_freq_sample1$beta3)/10000

quantile(hurricane_freq_sample1$beta0, c(0.05, 0.975)) 
quantile(hurricane_freq_sample1$beta1, c(0.05, 0.975))
quantile(hurricane_freq_sample1$beta2, c(0.05, 0.975))
quantile(hurricane_freq_sample1$beta3, c(0.05, 0.975))

mean(hurricane_freq_sample1$beta0)
mean(hurricane_freq_sample1$beta1)
mean(hurricane_freq_sample1$beta2)
mean(hurricane_freq_sample1$beta3)
```

Prediction
```{r}
preds = rep(0, 29)
for (i in 1:29){
  preds[i] <- mean(hurricane_freq_sample1$beta0) + mean(hurricane_freq_sample1$beta1)*hurricanes_per_year_clean$temp[i] + mean(hurricane_freq_sample1$beta2)*hurricanes_per_year_clean$avg_CO2[i] + mean(hurricane_freq_sample1$beta3)*hurricanes_per_year_clean$avg_NAO[i] 
}
pred2plot <- data.frame(cbind(hurricanes_per_year_clean$YEAR, preds, hurricanes_per_year_clean$TOTAL_H))
colnames(pred2plot) <- c("YEAR", "pred", "actual")

ggplot() +
  geom_point(data = pred2plot, aes(x = YEAR, y = pred), color = "blue") +
  geom_point(data = pred2plot, aes(x = YEAR, y = actual), color = "red")
```




### Model 2: Poisson regression

Question 1: Can we model the change in frequency of hurricanes based on trends in hurricane windspeed and category, temperature, CO2, and NAO index?

\

Format the data
```{r}
# table of locations where wind speed in max.  in cases where hurricane has multiple obervations at max wind speed, the center of these list locations is given as location of max wind speed
location_max <- hurricane_data %>%
  filter(YEAR > 1949) %>%
  group_by(YEAR, NAME) %>%
  filter(WIND_KTS == max(WIND_KTS)) %>%
  summarise(CENTER_LAT = mean(LAT), CENTER_LONG = mean(LONG), WIND_KTS = mean(WIND_KTS)) 
```


Build the model--Poisson regression (frequency of hurricanes)
```{r}
library(rjags)

head(location_max)
head(hurricanes_per_year_clean)

hur_mod <- " model {
  for (i in 1:length(TOTAL_H)) {
      TOTAL_H[i] ~ dpois(lam[i])
      log(lam[i]) = beta0 + beta1*X1[i] + beta2*X2[i] + beta3*X3[i]
  }

  beta0 ~ dnorm(0.0, 1.0/1e6)
  beta1 ~ dnorm(0.0, 1.0/1e4)
  beta2 ~ dnorm(0.0, 1.0/1e4)
  beta3 ~ dnorm(0.0, 1.0/1e4)
} "

# set up an algorithm to simulate the posterior by combining the model and data (x)
# set the random number seed

data_jags = as.list(hurricanes_per_year_clean[,2:5])
str(data_jags)

freq_jags <- jags.model(textConnection(hur_mod),data=list(TOTAL_H = hurricanes_per_year_clean$TOTAL_H, X1 = hurricanes_per_year_clean$temp, X2 = hurricanes_per_year_clean$avg_CO2,X3 = hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=2000))

# simulate a sample from the posterior
# note that we specify both mu and tau variables
freq_sim <- coda.samples(freq_jags, variable.names = c("beta0", "beta1", "beta2", "beta3"), n.iter=50000)

# store the samples in a data frame:
freq_sample <- data.frame(step = 1:50000, freq_sim[[1]])
head(freq_sample)
```

Looking at convergence
```{r}
library(MacBayes)
running_mean_plot(x=freq_sample$beta0, se=TRUE)
running_mean_plot(x=freq_sample$beta1, se=TRUE)
running_mean_plot(x=freq_sample$beta2, se=TRUE)
running_mean_plot(x=freq_sample$beta3, se=TRUE)
```

<<<<<<< HEAD
### Question 2.
=======
Testing
```{r}
exp(quantile(freq_sample$beta1, c(0.05, 0.975)))
exp(quantile(freq_sample$beta2, c(0.05, 0.975)))
exp(quantile(freq_sample$beta3, c(0.05, 0.975)))

mean(freq_sample$beta0)
mean(freq_sample$beta1)
mean(freq_sample$beta2)
mean(freq_sample$beta3)
```
>>>>>>> a22f559e9cd25af3504a9da11315b346e4d5b40e

```{r}
preds1 = rep(0, 29)
for (i in 1:29){
  preds1[i] <- exp(mean(freq_sample$beta0) + mean(freq_sample$beta1)*hurricanes_per_year_clean$temp[i] + mean(freq_sample$beta2)*hurricanes_per_year_clean$avg_CO2[i] + mean(freq_sample$beta3)*hurricanes_per_year_clean$avg_NAO[i] )
}

predPlot <- data.frame(cbind(hurricanes_per_year_clean$YEAR, preds1, hurricanes_per_year_clean$TOTAL_H))
colnames(predPlot) <- c("YEAR", "pred", "actual")

ggplot() +
  geom_point(data = predPlot, aes(x = YEAR, y = pred), color = "blue") +
  geom_point(data = predPlot, aes(x = YEAR, y = actual), color = "red")

```



## Question 2.

Can we model  the trend of average durations of hurricanes over the years?

\

Create new dataset on average durations per year
<<<<<<< HEAD
```{r echo=FALSE}
=======

```{r}
>>>>>>> a22f559e9cd25af3504a9da11315b346e4d5b40e
# Hurricane data only looking at Date and name
hurricane_time<-subset(recent_hurricanes, select = c(NAME,YEAR,MONTH,DAY))

#Group data by year and name and subtracting starting time from end time for each hurricane (duration)
foo <- hurricane_time %>%
  group_by(YEAR, NAME) %>%
  mutate(date = as.Date(paste(YEAR, MONTH, DAY, sep='-')), "%Y-%m-%d") %>%
  summarise(duration = max(date) - min(date))

#removed outlier
foo<-foo[!(foo$YEAR==1954 & foo$NAME=="ALICE"|foo$YEAR==2005),]

# Averaging duration per year
foo1<-group_by(foo, YEAR)%>%
  summarise(mean(duration))
colnames(foo1)[2] <- "Average_Duration"
ggplot(foo1, aes(x=Average_Duration))+
  geom_histogram(boundary=0,color="white", aes(y=..density..))
ggplot(foo1,aes(x=YEAR, y=Average_Duration))+
  geom_line()
```


### Model 1:

Question 2: Can we model  the trend of average durations of hurricanes over the years?

\

Prior: 
$$X \sim Norm(\mu,\tau)$$

Hyperpriors:
$$\mu \sim N(0, 1000)$$

$$\tau \sim \Gamma(2, 0.05)$$

Load the data
```{r}
#Group data by year and name and subtracting starting time from end time for each hurricane (duration)
foo <- hurricane_time %>%
  group_by(YEAR, NAME) %>%
  mutate(date = as.Date(paste(YEAR, MONTH, DAY, sep='-')), "%Y-%m-%d") %>%
  summarise(duration = max(date) - min(date))
#removed outlier
foo<-foo[!(foo$YEAR==1954 & foo$NAME=="ALICE"),]
# Averaging duration per year
foo<-group_by(foo, YEAR)%>%
  summarise(mean(duration))
```


Build the model
```{r}
#specify the model
duration_model <- "model{
    #Data
    for(i in 1:length(x)) {
        y[i] ~ dgamma(a[i],b)
        a[i] ~ dpois(lambda[i])
    }

    #Priors
    b ~ dgamma(1,1)
    beta0 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta1 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta2 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta3 ~ dnorm(0, 1/(1000)^2) #PRECISION
    tau ~ dgamma(0.001, 0.001)
}"

<<<<<<< HEAD


duration_jags <- jags.model(textConnection(duration_model), data=list(y=foo_train$Average_Duration, x1=hurricanes_per_year_clean$temp, x2=hurricanes_per_year_clean$avg_CO2, x3=hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))



duration_sim <- coda.samples(duration_jags, variable.names=c("b","beta0","beta1","beta2","beta3", "tau"), n.iter=10000)
=======
duration_jags <- jags.model(textConnection(duration_model), data=list(x=foo$Average_Duration), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

duration_sim <- coda.samples(duration_jags, variable.names=c("mu","tau", "theta"), n.iter=10000)
>>>>>>> a22f559e9cd25af3504a9da11315b346e4d5b40e

duration_samples <- data.frame(step=1:10000, duration_sim[[1]])
head(duration_samples)
plot(duration_sim)
summary(duration_sim)
```



### Model 2:

Question 2: Can we model  the trend of average durations of hurricanes over the years?

\

Prior: $$X \sim N(mu, tau)$$

Given length(x)= number of years
$$\mu\sim N(6.37, 10^2)$$

$$\tau \sim \Gamma(9, 6)$$

Create training and testing data
```{r message=FALSE, warning=FALSE}
data(foo)

## 50% of the sample size
smp_size <- floor(0.50 * nrow(foo))

## set the seed to make your partition reproductible
set.seed(123)
train_ind <- sample(seq_len(nrow(foo)), size = smp_size)

foo_train <- foo[train_ind, ]
foo_test <- foo[-train_ind, ]
```

Create model using training data
```{r}

library(rjags)

#specify the model
duration2_model <- "model{
    #Data
    for(i in 1:length(x)) {
        y[i] ~ dnorm(beta0 +beta1*x[i],tau)
    }

    #Priors
    beta0 ~ dnorm(6,1/1000)
    beta1~dnorm(0,1/1000)
    tau ~ dgamma(9, 6)
}"



duration2_jags <- jags.model(textConnection(duration2_model), data=list(x=foo_train$Average_Duration), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))



duration2_sim <- coda.samples(duration2_jags, variable.names=c("beta0","beta1","tau"), n.iter=10000)

duration2_samples <- data.frame(step=1:10000, duration_sim[[1]])
head(duration2_samples)
plot(duration2_sim)
summary(duration2_sim)

```


Test model on testing data
```{r}

library(rjags)

#specify the model
duration3_model <- "model{
    #Data
    for(i in 1:length(x)) {
        y[i] ~ dnorm(beta0 +beta1*x[i],tau)
    }

    #Priors
    beta0 ~ dnorm(6,1/1000)
    beta1~dnorm(0,1/1000)
    tau ~ dgamma(9, 6)
}"



duration3_jags <- jags.model(textConnection(duration3_model), data=list(x=foo_test$Average_Duration), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))



duration3_sim <- coda.samples(duration3_jags, variable.names=c("beta0","beta1","tau"), n.iter=10000)

duration3_samples <- data.frame(step=1:10000, duration_sim[[1]])
head(duration3_samples)
plot(duration3_sim)
summary(duration3_sim)

```


## Question 3:

Can we predict windspeed of hurricanes by NAO index, CO2, and temperature?

### Data
```{r}

# Create new data set with windspeed and year
windspeed_yr <- hurricane_data %>%
  group_by(YEAR) %>%
  summarise(avg_windspeed = mean(WIND_KTS))

# Merge this with the other data set
HURRICANES <- merge(x = hurricanes_per_year_clean, y = windspeed_yr, by = "YEAR", all.x = TRUE)

```

### Plots
```{r}

ggplot(HURRICANES, aes(x=avg_windspeed))+
  geom_histogram(boundary=0,color="white", aes(y=..density..))
ggplot(HURRICANES,aes(x=YEAR, y=avg_windspeed))+
  geom_line()

lo <- loess(HURRICANES$avg_windspeed~HURRICANES$YEAR)
plot(HURRICANES$YEAR,HURRICANES$avg_windspeed)
lines(predict(lo), col='red', lwd=2)

```

### Means
```{r}

mean(HURRICANES$avg_CO2)
mean(HURRICANES$temp)
mean(HURRICANES$avg_NAO)

var(HURRICANES$avg_CO2)
var(HURRICANES$temp)
var(HURRICANES$avg_NAO)

```

### Model


$Y_{i} =$ average windspeed of hurricane in year $i$

$X_{1} =$ average C02 in year $i$

$X_{2} =$ average temperature in year $i$

$X_{3} =$ average NAO in year $i$

We propose the following model:

$$Y_i | \beta_{0}, \beta_{1}, \beta_{2}, \beta_{3} \sim N(\beta_0 + \beta_{1}X_{1t} + \beta_{2}X_{2t} + \beta_{3}X_{3t}, \theta^2)$$

$$\beta_0 \sim N(0, 1000^2)$$

$$\beta_{1} \sim N(0, 1000^2)$$

$$\beta_{2} \sim N(0, 1000^2)$$

$$\beta_{3} \sim N(0, 1000^2)$$

$$(\theta^2)^{-1} \sim Gamma(0.001, 0.001)$$

```{r}

#*specify the model
library(rjags)

windSpeed_model <- "model{
    #Data
    for(i in 1:length(y)) {
        y[i] ~ dnorm(beta0 + beta1*x1[i] + beta2*x2[i] + beta3*x3[i], tau)

    }
    #Priors
    beta0 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta1 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta2 ~ dnorm(0, 1/(1000)^2) #PRECISION
    beta3 ~ dnorm(0, 1/(1000)^2) #PRECISION
    tau ~ dgamma(0.001, 0.001)

}"

#*set up an algorithm to simulate the posterior by
#*combining the model (games_model) and data (x)
#*set the random number seed
windSpeed_jags <- jags.model(textConnection(windSpeed_model),data=list(y=HURRICANES$avg_windspeed,x1=HURRICANES$avg_CO2, x2=HURRICANES$temp, x3=HURRICANES$avg_NAO),
                    inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

#*simulate a sample from the posterior
#*note that we specify both mu and tau variables
windSpeed_sim <- coda.samples(windSpeed_jags, variable.names=c("beta0","beta1","beta2","beta3"), n.iter=150000)


#*store the samples in a data frame:
windSpeed_sample <- data.frame(step=1:150000, windSpeed_sim[[1]])
head(windSpeed_sample)

```

```{r}

plot(windSpeed_sim)
summary(windSpeed_sim)

```

```{r}

sum(windSpeed_sample$beta0)/10000
sum(windSpeed_sample$beta1)/10000
sum(windSpeed_sample$beta2)/10000
sum(windSpeed_sample$beta3)/10000

quantile(windSpeed_sample$beta0, c(0.05, 0.975)) 
quantile(windSpeed_sample$beta1, c(0.05, 0.975))
quantile(windSpeed_sample$beta2, c(0.05, 0.975))
quantile(windSpeed_sample$beta3, c(0.05, 0.975))

```

The posterior mean $E(β0|y) = 103.8736$
The posterior mean $E(β1|y) = -0.1505875$
The posterior mean $E(β2|y) = 3.911151$
The posterior mean $E(β3|y) = 0.1622669$

The posterior 95% credible interval for $β0 = (45.61848, 157.49618)$
The posterior 95% credible interval for $β1 = (-0.2970584, 0.0372634)$
The posterior 95% credible interval for $β2 = (-5.460935, 14.519758)$
The posterior 95% credible interval for $β3 = (-2.451443, 3.319495)$

```{r}

running_mean_plot(x = windSpeed_sample$beta1, se = TRUE)
running_mean_plot(x = windSpeed_sample$beta2, se = TRUE)
running_mean_plot(x = windSpeed_sample$beta3, se = TRUE)

```


