# Project Checkpoint 6

**NAMES:**  Madeline Abbott, Aidan Teppema, Daisy Cho

\
\
\
\


```{r include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(MacBayes)
```

```{r include=FALSE}
# Loading the Data

# Hurricane Frequency
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")

hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ] 

# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))


# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)

# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))

# rename the columns
colnames(tempN) <- c("YEAR", "temp")

# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")

# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)

# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))

# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")

# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))

NAO_index2 <- NAO_index1[, -c(14)]

NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
  
NAO_index <- NAO_index3[-nrow(NAO_index3),] 

# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 

# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"

NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)

# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))

# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")

# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)
```

# Progress Made:

This week, we worked to fix the model that was not functioning last week (modeling windspeed).  We created a new model for annual hurricane rate using poisson regression.  We summarized this new model (and also last week's normal-normal model of hurricane rate) with running mean plots and found the credible intervals of the parameters.

# Group Member Roles:

* Madeline--summarize normal-normal model results and worked on poisson regression for hurricane rate.
* Daisy--
* Aidan--


# Modeling:

\

Things to think about:

* limit data to hurricanes in the Atlantic


## Question 1.


Can we model the change in frequency of hurricanes based on trends in hurricane windspeed and category, temperature, CO2, and NAO index?


\

Some plots...

\

Plot histogram of annual hurricane frequency
```{r}
# hurricanes
ggplot(hurricanes_per_year_clean, aes(x = TOTAL_H)) + geom_histogram(color = 'white', fill = 'goldenrod1', aes(y = ..density..))

# temp
ggplot(hurricanes_per_year_clean, aes(x = temp)) + geom_histogram(color = 'white', fill = 'coral', aes(y = ..density..))

#co2
ggplot(hurricanes_per_year_clean, aes(x = avg_CO2)) + geom_histogram(color = 'white', fill = 'darkorchid2', aes(y = ..density..))

# NAO
ggplot(hurricanes_per_year_clean, aes(x = avg_NAO)) + geom_histogram(color = 'white', fill = 'darkseagreen', aes(y = ..density..))
```

Plot distribution of variable per year
```{r}
#hurricanes
ggplot(hurricanes_per_year_clean, aes(x = YEAR, y = TOTAL_H)) + geom_point(color = "goldenrod1")

# temp
ggplot(hurricanes_per_year_clean, aes(x = YEAR, y = temp)) + geom_point(color = "coral")

# co2
ggplot(hurricanes_per_year_clean, aes(x = YEAR, y = avg_CO2)) + geom_point(color = "darkorchid2")

# NAO
ggplot(hurricanes_per_year_clean, aes(x = YEAR, y = avg_NAO)) + geom_point(color = "darkseagreen")
```

Plot hurricanes per year by temperature
```{r}
ggplot(hurricanes_per_year_clean, aes(x = temp, y = TOTAL_H)) + geom_point(color = "firebrick1")
```

Hurricanes per year by CO2 levels
```{r}
ggplot(hurricanes_per_year_clean, aes(x = avg_CO2, y = TOTAL_H)) + geom_point(color = "darkslateblue")
```

Hurricanes per year by CO2 levels
```{r}
ggplot(hurricanes_per_year_clean, aes(x = avg_NAO, y = TOTAL_H)) + geom_point(color = "aquamarine4")
```


### Model 1: Normal-normal

Let:

$Y_{t} =$ total hurricanes in year $t$

$X_{1t} =$ average yearly temperature in year $t$

$X_{2t} =$ yearly CO2 concentration in year $t$

$X_{3t} =$ NAO index in year $t$

We propose the following model:

$$Y_t | \beta_{0}, \beta_{1}, \beta_{2}, \beta_{3} \sim N(\beta_0 + \beta_{1}X_{1t} + \beta_{2}X_{2t} + \beta_{3}X_{3t}, \theta^2)$$

$$\beta_0 \sim N(0, 1000^2)$$

$$\beta_{1} \sim N(0, 1000^2)$$

$$\beta_{2} \sim N(0, 1000^2)$$

$$\beta_{3} \sim N(0, 1000^2)$$

$$(\theta^2)^{-1} \sim Gamma(0.001, 0.001)$$


Build the model--normal-normal
```{r}
# specify the model
hurricane_freq_model1 <- "model{
  # Data
  for (t in 1:length(y)) {
    y[t] ~ dnorm(beta0 + beta1*x1[t] + beta2*x2[t] + beta3*x3[t], tau) #PRECISION
    }

  #Priors
  beta0 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta1 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta2 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta3 ~ dnorm(0, 1/(1000)^2) #PRECISION
  tau ~ dgamma(0.001, 0.001)
}"


# set up an algorithm to simulate the posterior by combining the model (hurricane_freq_model1) and data (y)
# set the random number seed
hurricane_freq_jags1 <- jags.model(textConnection(hurricane_freq_model1), data=list(y=hurricanes_per_year_clean$TOTAL_H, x1=hurricanes_per_year_clean$temp, x2=hurricanes_per_year_clean$avg_CO2, x3=hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=2000))

# simulate a sample from the posterior
# note that we specify both mu and tau variables
hurricane_freq_sim1 <- coda.samples(hurricane_freq_jags1, variable.names = c("beta0", "beta1", "beta2", "beta3", "tau"), n.iter=50000)

# store the samples in a data frame:
hurricane_freq_sample1 <- data.frame(step = 1:50000, hurricane_freq_sim1[[1]])
head(hurricane_freq_sample1)

summary(hurricane_freq_sim1)
plot(hurricane_freq_sim1)
```


Do parameter estimations converge?
```{r}
running_mean_plot(x = hurricane_freq_sample1$beta1, se = TRUE)
running_mean_plot(x = hurricane_freq_sample1$beta2, se = TRUE)
running_mean_plot(x = hurricane_freq_sample1$beta3, se = TRUE)
```

Are the predictor variables useful?  Look at model parameters
```{r}
quantile(hurricane_freq_sample1$beta1, c(0.05, 0.975))
quantile(hurricane_freq_sample1$beta2, c(0.05, 0.975))
quantile(hurricane_freq_sample1$beta3, c(0.05, 0.975))

mean(hurricane_freq_sample1$beta0)
mean(hurricane_freq_sample1$beta1)
mean(hurricane_freq_sample1$beta2)
mean(hurricane_freq_sample1$beta3)
```

Prediction
```{r}
preds = rep(0, 29)
for (i in 1:29){
  preds[i] <- mean(hurricane_freq_sample1$beta0) + mean(hurricane_freq_sample1$beta1)*hurricanes_per_year_clean$temp[i] + mean(hurricane_freq_sample1$beta2)*hurricanes_per_year_clean$avg_CO2[i] + mean(hurricane_freq_sample1$beta3)*hurricanes_per_year_clean$avg_NAO[i] 
}
pred2plot <- data.frame(cbind(hurricanes_per_year_clean$YEAR, preds, hurricanes_per_year_clean$TOTAL_H))
colnames(pred2plot) <- c("YEAR", "pred", "actual")

ggplot() +
  geom_point(data = pred2plot, aes(x = YEAR, y = pred), color = "blue") +
  geom_point(data = pred2plot, aes(x = YEAR, y = actual), color = "red")
```




### Model 2: Poisson regression

(hurricane frequency) 

\

Format the data
```{r}
# table of locations where wind speed in max.  in cases where hurricane has multiple obervations at max wind speed, the center of these list locations is given as location of max wind speed
location_max <- hurricane_data %>%
  filter(YEAR > 1949) %>%
  group_by(YEAR, NAME) %>%
  filter(WIND_KTS == max(WIND_KTS)) %>%
  summarise(CENTER_LAT = mean(LAT), CENTER_LONG = mean(LONG), WIND_KTS = mean(WIND_KTS)) 
```


Build the model--Poisson regression (frequency of hurricanes)
```{r}
library(rjags)

head(location_max)
head(hurricanes_per_year_clean)

hur_mod <- " model {
  for (i in 1:length(TOTAL_H)) {
      TOTAL_H[i] ~ dpois(lam[i])
      log(lam[i]) = beta0 + beta1*X1[i] + beta2*X2[i] + beta3*X3[i]
  }

  beta0 ~ dnorm(0.0, 1.0/1e6)
  beta1 ~ dnorm(0.0, 1.0/1e4)
  beta2 ~ dnorm(0.0, 1.0/1e4)
  beta3 ~ dnorm(0.0, 1.0/1e4)
} "

# set up an algorithm to simulate the posterior by combining the model and data (x)
# set the random number seed

data_jags = as.list(hurricanes_per_year_clean[,2:5])
str(data_jags)

freq_jags <- jags.model(textConnection(hur_mod),data=list(TOTAL_H = hurricanes_per_year_clean$TOTAL_H, X1 = hurricanes_per_year_clean$temp, X2 = hurricanes_per_year_clean$avg_CO2,X3 = hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=2000))

# simulate a sample from the posterior
# note that we specify both mu and tau variables
freq_sim <- coda.samples(freq_jags, variable.names = c("beta0", "beta1", "beta2", "beta3"), n.iter=50000)

# store the samples in a data frame:
freq_sample <- data.frame(step = 1:50000, freq_sim[[1]])
head(freq_sample)
```

Do model parameters converge?
```{r}
running_mean_plot(x=freq_sample$beta0, se=TRUE)
running_mean_plot(x=freq_sample$beta1, se=TRUE)
running_mean_plot(x=freq_sample$beta2, se=TRUE)
```



Are the predictor variables useful?  Look at model parameters
```{r}
quantile(freq_sample$beta1, c(0.05, 0.975))
quantile(freq_sample$beta2, c(0.05, 0.975))
quantile(freq_sample$beta3, c(0.05, 0.975))

mean(freq_sample$beta0)
mean(freq_sample$beta1)
mean(freq_sample$beta2)
mean(freq_sample$beta3)
```




