# Project Checkpoint 5

**NAMES:**  Madeline Abbott, Aidan Teppema, Daisy Cho

\
\
\
\


```{r include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
#library(shiny)
library(rjags)
library(naivebayes)
```

# Progress Made:

We formulated 3 potential, relevant research questions. We designed models for these and created rjags simulations to simulate these models.


# Group Member Roles:

* Madeline--questions 2: constructed simple models for predicting hurricane frequency based on NOA index, CO2 levels, and temperature trends.
* Daisy--constructed 2 models for research question one to model durations of hurricanes over the years.
* Aidan--worked on merging wind speed data and designed models for research question 3

# Research questions:

### 1.
Can we model the durations of hurricanes over the years?

Model 1:

Prior: $X\sim Norm(\mu,\tau)$

Hyperpriors:
$\mu\sim (0, 1000)$

$\tau \sim (2, 0.05)$

Model 2:

Prior: $X\sim Norm(mu, tau$

Given length(x)= number of years
$\mu\sim (6.37, 10^2)$

$\tau \sim (9, 6)$

### 2.

Can we predict the severity of these disasters based on trends in hurricane wind speed, temperature, C02, and NAO index? (figure out categories maybe select= c(categories 1-5))

### 3.

Can we predict the intensity of wind speed of these disasters based on trends in average yearly temperature, average yearly C02, and yearly NAO index?

## Building simple models


```{r include=FALSE}
# Loading the Data

# Hurricane Frequency
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")

hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

# select only hurricanes in 1950 and after
recent_hurricanes <- hurricane_data[hurricane_data$YEAR >= 1950, ]

# table of hurricanes per year
hurricanes_per_year <- recent_hurricanes %>%
  group_by(YEAR) %>%
  summarise(TOTAL_H = n_distinct(NAME))


# Temperature Data
temp_data <- read.csv("ZonAnn.csv")
temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)

# select only northern hemisphere temp data
tempN <- subset(temp_recent, select = c(Year, NHem))

# rename the columns
colnames(tempN) <- c("YEAR", "temp")

# join temperature to hurricanes per year table
hurricanes_per_year2 <- merge(x = hurricanes_per_year, y = tempN, by = "YEAR", all.x = TRUE)

# CO2 Data
CO2 <- read_csv("CO2.csv")

# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)

# get average by year
CO2_yr <- CO2 %>%
  group_by(year) %>%
  summarise(avg_CO2 = mean(average))

# rename columns
colnames(CO2_yr) <- c("YEAR", "avg_CO2")

# join with hurricane per year table
hurricanes_per_year3 <- merge(x = hurricanes_per_year2, y = CO2_yr, by = "YEAR", all.x = TRUE)
# FIX THIS (bc only 1980 onward available)

# NAO Index Data
NAO_index1 <- read_csv("NAO_index_monthly.csv",
    col_types = cols(Apr = col_number(),
        Aug = col_number(), Dec = col_number(),
        Feb = col_number(), Jan = col_number(),
        Jul = col_number(), Jun = col_number(),
        Mar = col_number(), May = col_number(),
        Nov = col_number(), Oct = col_number(),
        Sep = col_number()))

NAO_index2 <- NAO_index1[, -c(14)]

NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))

NAO_index <- NAO_index3[-nrow(NAO_index3),]

# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1)

# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"

NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)

# group by year
NAO_yr <- NAO_index %>%
  group_by(year) %>%
  summarise(avg_NAO = mean(index))

# rename columns
colnames(NAO_yr) <- c("YEAR", "avg_NAO")

# join with hurricane data
hurricanes_per_year4 <- merge(x = hurricanes_per_year3, y = NAO_yr, by = "YEAR", all.x = TRUE)

# Remove all rows with missing data
hurricanes_per_year_clean<-na.omit(hurricanes_per_year4)
```





## 1.
```{r}
hurricanes_recent<-recent_hurricanes[!(recent_hurricanes$NAME=="NOTNAMED"),]

# Hurricane data only looking at Date and name
hurricane_time<-subset(hurricanes_recent, select = c(NAME,YEAR,MONTH,DAY))
#Group data by year and name and subtracting starting time from end time for each hurricane (duration)
foo <- hurricane_time %>%
  group_by(YEAR, NAME) %>%
  mutate(date = as.Date(paste(YEAR, MONTH, DAY, sep='-')), "%Y-%m-%d") %>%
  summarise(duration = max(date) - min(date))
#removed outlier
foo<-foo[!(foo$YEAR==1954 & foo$NAME=="ALICE"),]
# Averaging duration per year
foo<-group_by(foo, YEAR)%>%
  summarise(mean(duration))
colnames(foo)[2] <- "Average_Duration"
ggplot(foo, aes(x=Average_Duration))+
  geom_histogram(boundary=0,color="white", aes(y=..density..))
ggplot(foo,aes(x=YEAR, y=Average_Duration))+
  geom_line()
```

```{r}
library(rjags)

#specify the model
duration_model <- "model{
    #Data
    for(i in 1:length(x)) {
        x[i] ~ dnorm(mu,tau)
    }

    #Priors
    mu ~ dnorm(0,1/1000)
    tau ~ dgamma(2, 0.05)
}"



duration_jags <- jags.model(textConnection(duration_model), data=list(x=foo$Average_Duration), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))



duration_sim <- coda.samples(duration_jags, variable.names=c("mu","tau", "theta"), n.iter=10000)

duration_samples <- data.frame(step=1:10000, duration_sim[[1]])
head(duration_samples)
```

```{r}
#specify the model
duration2_model <- "model{
    #Data
    for(i in 1:length(x)) {
        x[i] ~ dnorm(mu,tau)
    }

    #Priors
    mu ~ dnorm(6.37,1/10)
    tau ~ dgamma(9, 6)
}"

duration2_jags <- jags.model(textConnection(duration2_model), data=list(x=foo$Average_Duration), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

duration2_sim <- coda.samples(duration2_jags, variable.names=c("mu","tau"), n.iter=10000)

duration2_samples <- data.frame(step=1:10000, duration2_sim[[1]])
head(duration2_samples)
```


### 2.

Can we model the change in frequency of hurricanes based on trends in hurricane windspeed and category, temperature, CO2, and NAO index?


Plot distribution of hurricanes per year
```{r}
ggplot(hurricanes_per_year, aes(x = YEAR, y = TOTAL_H)) + geom_point(color = "orange")
```

Plot hurricanes per year by temperature
```{r}
ggplot(hurricanes_per_year_clean, aes(x = temp, y = TOTAL_H)) + geom_point(color = "skyblue")
```

Hurricanes per year by CO2 levels
```{r}
ggplot(hurricanes_per_year_clean, aes(x = avg_CO2, y = TOTAL_H)) + geom_point(color = "darkslateblue")
```

Hurricanes per year by CO2 levels
```{r}
ggplot(hurricanes_per_year_clean, aes(x = avg_NAO, y = TOTAL_H)) + geom_point(color = "darkseagreen")
```


### Model 1:

Let:

$Y_{t} =$ total hurricanes in year $t$

$X_{1t} =$ average yearly temperature in year $t$

$X_{2t} =$ yearly CO2 concentration in year $t$

$X_{3t} =$ NAO index in year $t$

We propose the following model:

$$Y_t | \beta_{0}, \beta_{1}, \beta_{2}, \beta_{3} \sim N(\beta_0 + \beta_{1}X_{1t} + \beta_{2}X_{2t} + \beta_{3}X_{3t}, \theta^2)$$

$$\beta_0 \sim N(0, 1000^2)$$

$$\beta_{1} \sim N(0, 1000^2)$$

$$\beta_{2} \sim N(0, 1000^2)$$

$$\beta_{3} \sim N(0, 1000^2)$$

$$(\theta^2)^{-1} \sim Gamma(0.001, 0.001)$$


```{r}
# specify the model
hurricane_freq_model1 <- "model{
  # Data
  for (t in 1:length(y)) {
    y[t] ~ dnorm(beta0 + beta1*x1[t] + beta2*x2[t] + beta3*x3[t], tau) #PRECISION
    }

  #Priors
  beta0 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta1 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta2 ~ dnorm(0, 1/(1000)^2) #PRECISION
  beta3 ~ dnorm(0, 1/(1000)^2) #PRECISION
  tau ~ dgamma(0.001, 0.001)
}"


# set up an algorithm to simulate the posterior by combining the model (hurricane_freq_model1) and data (y)
# set the random number seed
hurricane_freq_jags1 <- jags.model(textConnection(hurricane_freq_model1), data=list(y=hurricanes_per_year_clean$TOTAL_H, x1=hurricanes_per_year_clean$temp, x2=hurricanes_per_year_clean$avg_CO2, x3=hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

# simulate a sample from the posterior
# note that we specify both mu and tau variables
hurricane_freq_sim1 <- coda.samples(hurricane_freq_jags1, variable.names = c("beta0", "beta1", "beta2", "beta3", "tau"), n.iter=100)

# store the samples in a data frame:
hurricane_freq_sample1 <- data.frame(step = 1:100, hurricane_freq_sim1[[1]])
head(hurricane_freq_sample1)

summary(hurricane_freq_sample1)
#plot(hurricane_freq_sample1)
```



### Model 2:

Let:

$Y_{t} =$ total hurricanes in year $t$

$X_{1t} =$ average yearly temperature in year $t$

$X_{2t} =$ yearly CO2 concentration in year $t$

$X_{3t} =$ NAO index in year $t$

We propose the following model:

$$Y_t | \beta_{0}, \beta_{1}, \beta_{2}, \beta_{3} \sim N(\beta_0 + \beta_{1}X_{1t} + \beta_{2}X_{2t} + \beta_{3}X_{3t}, \theta^2)$$

$$\beta_0 \sim N(\mu_{0}, 1000^2)$$

$$\beta_{1t} \sim N(\mu_{1}, 1000^2)$$

$$\beta_{2t} \sim N(\mu_{2}, 1000^2)$$

$$\beta_{3t} \sim N(\mu_{3}, 1000^2)$$

$$(\theta^2)^{-1} \sim Gamma(0.001, 0.001)$$

$$\mu_{0} \sim N(0, 1000^2)$$

$$\mu_{1} \sim N(0, 1000^2)$$

$$\mu_{2} \sim N(0, 1000^2)$$

$$\mu_{3} \sim N(0, 1000^2)$$

```{r}
# specify the model
hurricane_freq_model2 <- "model{
  # Data
  for (t in 1:length(y)) {
    y[t] ~ dnorm(beta0 + beta1*x1[t] + beta2*x2[t] + beta3*x3[t], tau) #PRECISION
    }

  #Priors
  beta0 ~ dnorm(mu0, 1/(1000)^2) #PRECISION
  beta1 ~ dnorm(mu1, 1/(1000)^2) #PRECISION
  beta2 ~ dnorm(mu2, 1/(1000)^2) #PRECISION
  beta3 ~ dnorm(mu3, 1/(1000)^2) #PRECISION
  tau ~ dgamma(0.001, 0.001)
  mu0 ~ dnorm(0, 1/(1000)^2)
  mu1 ~ dnorm(0, 1/(1000)^2)
  mu2 ~ dnorm(0, 1/(1000)^2)
  mu3 ~ dnorm(0, 1/(1000)^2)
}"


# set up an algorithm to simulate the posterior by combining the model (hurricane_freq_model2) and data (y)
# set the random number seed
hurricane_freq_jags2 <- jags.model(textConnection(hurricane_freq_model2), data=list(y=hurricanes_per_year_clean$TOTAL_H, x1=hurricanes_per_year_clean$temp, x2=hurricanes_per_year_clean$avg_CO2, x3=hurricanes_per_year_clean$avg_NAO), inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

# simulate a sample from the posterior
# note that we specify both mu and tau variables
hurricane_freq_sim2 <- coda.samples(hurricane_freq_jags2, variable.names = c("beta0", "beta1", "beta2", "beta3", "tau"), n.iter=100)

# store the samples in a data frame:
hurricane_freq_sample2 <- data.frame(step = 1:100, hurricane_freq_sim2[[1]])
head(hurricane_freq_sample2)

summary(hurricane_freq_sample2)
#plot(hurricane_freq_sample1)
```


## 3.

Can we predict the severity of these disasters based on trends in hurricane windspeed and category, temperature, C02, and NAO index?

This is an example rjags for this model. It doesn't run yet because of issues with data merging.


*specify the model
windSpeed_model <- "model{
    #Data
    for(i in 1:length(y)) {
        y[i] ~ dnorm(beta0 + beta1[i] + beta2[i] + beta3[i], tau)

    }
    #Priors
    beta1 ~ dnorm(0, 1/100) #PRECISION
    beta2 ~ dnorm(0, 1/100) #PRECISION
    beta3 ~ dnorm(0, 1/100) #PRECISION
    tau ~ dgamma(1,10)

}"

*set up an algorithm to simulate the posterior by
*combining the model (games_model) and data (x)
*set the random number seed
windSpeed_jags <- jags.model(textConnection(windSpeed_model), data=list(y=hurricane_data$WIND_KTS, beta1=HURRICANES$av_CO2, beta2=HURRICANES$temp,beta3=HURRICANES$avg_NAO),
                    inits=list(.RNG.name="base::Wichmann-Hill", .RNG.seed=1989))

*simulate a sample from the posterior
*note that we specify both mu and tau variables
windSpeed_sim <- coda.samples(windSpeed_jags, variable.names=c("beta0","beta1","beta2"), n.iter=10000)


*store the samples in a data frame:
windSpeed_sample <- data.frame(step=1:10000, BB_sim[[1]])
head(windSpeed_sample)