# Project Checkpoint 4

**NAMES:**  Madeline Abbott, Aidan Teppema, Daisy Cho

\
\
\
\


```{r include=FALSE}
library(readr)
library(dplyr)
library(tidyr)
library(ggplot2)
library(zoo)
library(lubridate)
library(ggmap)
library(reshape2)
library(MASS)
library(viridis)
```

# Progress Made:

We began our exploratory analysis of the data by creating plots of potential variables of interest (such as hurricane frequency, category frequency, maximum wind speed, NAO index, and CO2 levels) over time.  We looked for trends in these variables over time--trends that we could potentially use in our model.  This visuals help exlain the variables and we plan to adapt many of these plots to include them in our final report.  Maybe even in a shiny ap!


# Group Member Roles:

* Madeline--created some plots of hurricane data and NAO index data
* Daisy--created plots of temperature data
* Aidan--created plots of CO2 data


# Data

# Hurricanes
```{r}
hurricane_data <- read_csv("Historical_Tropical_Storm_Tracks.csv")

hurricane_data<-na.omit(hurricane_data)
hurricane_data<-subset(hurricane_data, select=-c(BTID, AD_TIME))

#transform(hurricane_data, char = as.numeric(LAT))
#transform(hurricane_data, char = as.numeric(LONG))

hurricane_data$LAT <- as.numeric(hurricane_data$LAT)
hurricane_data$LONG <- as.numeric(hurricane_data$LONG)

dim(hurricane_data)
names(hurricane_data)
head(hurricane_data)
summary(hurricane_data)
```

### Exploratory analysis
```{r}
# select only the hurricanes after 1950 (this is when the naming system started)
hurricanes_recent <- hurricane_data %>%
  filter(YEAR > 1949)
```

Some plots...
```{r}
# Plotting hurricane denisty by location

get_density <- function(x, y, n = 100) {
  dens <- MASS::kde2d(x = x, y = y, n = n)
  ix <- findInterval(x, dens$x)
  iy <- findInterval(y, dens$y)
  ii <- cbind(ix, iy)
  return(dens$z[ii])
}

hurricanes_recent$density <- get_density(hurricanes_recent$LONG, hurricanes_recent$LAT)
# creating a sample data.frame with your lat/lon points
hurricanes_recent$LAT <- as.numeric(hurricanes_recent$LAT)
hurricanes_recent$LONG <- as.numeric(hurricanes_recent$LONG)
#transform(hurricanes_recent, char = as.numeric(LAT))
#transform(hurricanes_recent, char = as.numeric(LONG))

# getting the map
map_hurricanes <- get_map(location = c(lon = mean(hurricanes_recent$LONG), lat = mean(hurricanes_recent$LAT)), zoom = 2, maptype = "terrain", scale = 2)

# plotting the map with some points on it
ggmap(map_hurricanes) +
  geom_point(data = hurricanes_recent, aes(x = LONG, y = LAT, alpha = 0.1, color = density)) + scale_color_viridis() + guides(size=FALSE)
```

```{r}
# Total number of hurricanes per year
hurricanes_per_year <- group_by(hurricanes_recent, YEAR) %>%
  summarize(length(unique(NAME)))
colnames(hurricanes_per_year)[2] <- "TOTAL_HURRICANES"
ggplot(hurricanes_per_year, aes(x = YEAR, y = TOTAL_HURRICANES)) + geom_point()
```

```{r}
# Total number of hurricanes of each category per year
hurricanes_categories <- group_by(hurricanes_recent, YEAR) %>%
  count(CAT)
colnames(hurricanes_categories)[3] <- "TOTAL_HURRICANES"
ggplot(hurricanes_categories, aes(x = YEAR, y = TOTAL_HURRICANES, color = CAT, fill = CAT)) + geom_bar(stat = "identity")
```

```{r fig.height = 2, fig.width = 8}
ggplot(hurricanes_categories, aes(x = YEAR, y = TOTAL_HURRICANES, color = CAT, fill = CAT)) + geom_bar(stat = "identity", position = position_dodge(width = 0.9))
```

```{r}
# Average pressure by year
hurricane_pressure <- group_by(hurricanes_recent, YEAR) %>%
  summarize(mean(PRESSURE))
colnames(hurricane_pressure)[2] <- 'AVG_PRESSURE'
colnames(hurricane_pressure)[1] <- "YEAR"
ggplot(hurricane_pressure, aes(x = YEAR, y = AVG_PRESSURE)) + geom_point() + ylab("Average pressure")
```


```{r}
# Average max wind speed by year
hurricanes_wind <- group_by(hurricanes_recent, YEAR) %>%
  summarize(mean(WIND_KTS))
colnames(hurricanes_wind)[2] <- "AVG_WIND"
ggplot(hurricanes_wind, aes(x = YEAR, y = AVG_WIND)) + geom_point() + ylab("Average wind speed (kts)")

# If you want to save a plot...
#ggsave("hurri_categories2.pdf", plot = last_plot(), height = 5, width = 10)
```


#Temperature
```{r}
temp_data <- read.csv("ZonAnn.csv")

temp_data[!complete.cases(temp_data),]
temp_data<-na.omit(temp_data)
temp_data<-temp_data[,-(5:15),drop=FALSE]
temp_recent <- temp_data %>%
  filter(Year > 1949)
dim(temp_recent)
names(temp_recent)
head(temp_recent)
summary(temp_recent)

# plot of Global, Northern Hemisphere, Southern Hemisphere deviation from corresponding mean by year
ggplot(temp_recent,aes(x=Year))+
  geom_line(aes(y=Glob, color="Global"))+
  geom_line(aes(y=NHem, color="Northern Hemisphere"))+
  geom_line(aes(y=SHem, color="Southern Hemisphere"))+
<<<<<<< HEAD
  ylab("Temperature Anomolies")



#ggplot(hurricanes_categories, aes(x = YEAR, y = TOTAL_HURRICANES, color = CAT, fill = CAT)) + 
  #geom_bar(stat = "identity") +
  #geom_abline(data=temp_data, aes(x=Year, y=Glob))
=======
  ylab("Temperature Anomolies")+
  geom_smooth(method='lm', aes(Year,Glob), colour="red", size=0.5, se=FALSE)+
  geom_smooth(method='lm', aes(Year,NHem),colour="green", size=0.5, se=FALSE)+
  geom_smooth(method='lm', aes(Year,SHem), colour="blue", size=0.5, se=FALSE)
>>>>>>> e790dea0cd7a3a3e049a57dff9a41681643a5d8a
```


#CO2

```{r}
<<<<<<< HEAD

# Show increasing CO2 over the years by month
library(ggplot2)
library(lubridate)

CO2 <- read_csv("~/Desktop/Bayes Capstone/CO2.csv")

ggplot(CO2, aes(x = month, y = average, color = year)) + geom_point()

```

# CO2
```{r}

=======
CO2 <- read_csv("CO2.csv")
>>>>>>> e790dea0cd7a3a3e049a57dff9a41681643a5d8a
# Alter CO2 to remove unnecessary variable trend
CO2 <- subset(CO2, select = -trend)

# Create subset of CO2 data from year 2000 to show only fluxuations in CO2 levels throughout one year

# Show increasing CO2 over the years by month
ggplot(CO2, aes(x = month, y = average, color = year)) + geom_point()


# Create subset of CO2 data from year 2000
# to show only fluxuations in CO2 levels throughout one year

CO2_2000 <- subset(CO2, year == 2000)
head(CO2_2000)

# Plot this example
plot(CO2_2000$average ~ CO2_2000$month)


# Create time series to show variation over months and years
Average <- ts(CO2$average, frequency = 12, start = 1980)


# Create time series to show variation over months and years
Average <- ts(CO2$average, frequency = 12, start = 1980)

# Plot and add axis labels
v1 <- c(1980,1985,1990,1995,2000,2005,2010,2015)
v2 <- c(1980,1985,1990,1995,2000,2005,2010,2015)
plot(Average)
axis(side = 1, 
     at = v1, 
     labels = v2,
     tck=-.05)

# Create colored time series 
CO2$Date <- as.yearmon(paste(CO2$year, CO2$month), "%Y %m")

# Plot this time series
ggplot(CO2, aes(Date, Average)) + geom_line( aes(colour = year))

# Create colored time series 
CO2$Date <- as.yearmon(paste(CO2$year, CO2$month), "%Y %m")

# Plot time series
ggplot(CO2, aes(Date, Average)) + geom_line( aes(colour = year))

# Add trend line
ggplot(CO2, aes(Date, Average)) + geom_line( aes(colour = year)) + geom_smooth()

# Show increasing CO2 over the years by month
ggplot(CO2, aes(x = month, y = average, color = year)) + geom_point()
```



# NAO Index Data
```{r}
NAO_index1 <- read_csv("NAO_index_monthly.csv", 
    col_types = cols(Apr = col_number(), 
        Aug = col_number(), Dec = col_number(), 
        Feb = col_number(), Jan = col_number(), 
        Jul = col_number(), Jun = col_number(), 
        Mar = col_number(), May = col_number(), 
        Nov = col_number(), Oct = col_number(), 
        Sep = col_number()))

NAO_index2 <- NAO_index1[, -c(14)]

NAO_index3 <- NAO_index2 %>%
  rename(year = X1) %>%
  mutate(Jul = replace(Jul, Jul < -50| Jul > 50, "")) %>%
  mutate(Aug = replace(Aug, Aug < -50| Aug > 50, "")) %>%
  mutate(Sep = replace(Sep, Sep < -50| Sep > 50, "")) %>%
  mutate(Oct = replace(Oct, Oct < -50| Oct > 50, "")) %>%
  mutate(Nov = replace(Nov, Nov < -50| Nov > 50, "")) %>%
  mutate(Dec = replace(Dec, Dec < -50| Dec > 50, ""))
  

NAO_index <- NAO_index3[-nrow(NAO_index3),] 

dim(NAO_index)
names(NAO_index)
head(NAO_index)
summary(NAO_index)


# remove 2017 (data for year is incomplete)
NAO_index <- head(NAO_index, -1) 

# reshape table
NAO_index <- melt(NAO_index, id=c("year"))
colnames(NAO_index)[2] <- "month"
colnames(NAO_index)[3] <- "index"

NAO_index$year <- as.factor(NAO_index$year)
NAO_index$index <- as.numeric(NAO_index$index)

# group by month
NAO_avgMonth <- aggregate(NAO_index[, 3], list(NAO_index$month), mean)

# plot
ggplot(data = NAO_avgMonth, aes(x = Group.1, y = x)) + geom_point(color = "blue") + xlab("month") + ylab("NAO index, averaged across all years")
```

